---
title: "cFIT"
output: html_document
date: "2023-08-24"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Getting started

First load the package

```{r, eval=T}
library(cFIT)
library(caret)
options(warn = -1)
data_file_path = "D:/OneDrive/OneDrive - UW-Madison/Kris/Code/curatedMetagenomicDataAnalyses/vignettes/"
```

Load dataset and deal with NAs
```{r}
crc_whole <- read.csv(paste0(data_file_path, "crc_whole_taxon.csv"))
# crc_sepa <- read.csv(paste0(data_file_path, "crc_separate_taxon.csv"))
crc_meta <- read.csv(paste0(data_file_path, "crc_meta.csv"))
study_names <- unique(crc_whole[,2])
study_nobs <- c(154,60,81,80,60,80,104,125,616,128,156)

# remove NA in the response
rows_with_na <- which(!complete.cases(crc_whole[,3])) #
crc_whole <- crc_whole[-rows_with_na, ]
crc_meta <- crc_meta[-rows_with_na, ]

# temperally set NA in the count matrix as 0
crc_whole[is.na(crc_whole)] <- 0

colnames(crc_whole) <- gsub("__", "-", colnames(crc_whole))
colnames(crc_whole) <- gsub("_", "-", colnames(crc_whole))

print(paste(nrow(crc_whole), 'observations, ', ncol(crc_whole)-3, 'species'))
```


```{r, eval=T}
meta_data = list(FengQ_2015=data.frame(tech=rep(name, times = study_nobs[1])),
                GuptaA_2019=data.frame(tech=rep(name, times = study_nobs[2])),
                HanniganGD_2017=data.frame(tech=rep(name, times = study_nobs[3])),
                ThomasAM_2018a=data.frame(tech=rep(name, times = study_nobs[4])),
                ThomasAM_2018b=data.frame(tech=rep(name, times = study_nobs[5])),
                ThomasAM_2019_c=data.frame(tech=rep(name, times = study_nobs[6])),
                VogtmannE_2016=data.frame(tech=rep(name, times = study_nobs[7])),
                WirbelJ_2018=data.frame(tech=rep(name, times = study_nobs[8])),
                YachidaS_2019=data.frame(tech=rep(name, times = study_nobs[9])),
                YuJ_2015=data.frame(tech=rep(name, times = study_nobs[10])),
                ZellerG_2014=data.frame(tech=rep(name, times = study_nobs[11])))
```


```{r, eval=T}
# extract the raw counts and metadata for data sets from 5 technologies
data.list = split_dataset_by_batch(X=as.matrix(crc_whole[,4:ncol(crc_whole)]), 
                                   batch = crc_whole[,2], 
                                   labels = crc_meta$study_condition,
                                   metadata = crc_meta, 
                                   dataset.name = 'crc:')

```

## Data integration

cFIT integration is composed of the following procedure:

* Gene selection. Select highly variable genes that are shared among data sources. Here we adopt the proposed procedure from [Seurat V3](https://satijalab.org/seurat/).
* Data preprocessing. Library size normalization and log transformation
* Model parameter estimation. The model parameters are estimated throught iterative NMF
* Post processing, evaluation, and downstream analysis.

Select 2000 highly variable genes
```{r, eval=T}
# select highly variable genes
species = select_genes(data.list$X.list, ngenes=1000, verbose=F)
```

Preprocess to normalize by library size and log transform
```{r,  eval=T}
# data preprocessing
exprs.list = preprocess_for_integration(data.list$X.list, species, scale.factor=10^4, scale=T, center=F)
# exprs.list is a list with number of studies' subsets (11 subsets)
```

We first integrate data sets from ten studies, and then transfer the learned factor matrix to a relatively small dataset: GuptaA_2019 (60 observations).

Perform data integration with `CFITIntegrate`. 
The only parameter to adjust is `r`, which is the number of factors of the common factor matrix (shared across datasets). The algorithm is relatively robust to `r` as long as r approximate the number of underlying distinct cell types across all data sets. 

Other minor parameters that can be adjusted is 
`max.niter` (100 by default), 
`tol` tolerance used in stopping criteria (1e-5 by default), 
`nrep` number of repeats of the run (1 by default, require much longer if increased), 
`init` parameter initialization (NULL). The parallel is control by `future` R package, by default `sequential` plan is used. To enable parallel computation, set `future.plan='multicore` and set the `workers` as the number of cores to use. Note that Rstudio does not support 'multicore'.

```{r, eval=T}
# integrate four largest datasets, takes roughly 20 minutes to run without parallization
ref.data = c(1,3,4,5,6,7,8,9,10,11)
int.out = CFITIntegrate(X.list=exprs.list[ref.data], r=15, max.niter=100, future.plan='sequential', seed=0, verbose=F)

# For large datasets we recommende using the sketched version for fast convergence.
# int.out = CFITIntegrate_sketched(X.list=exprs.list[ref.data], r=15, subsample.prop = 0.1,
#                                  max.niter=100, early.stopping=10, tol=1e-8, 
#                                  future.plan='sequential', seed=42, verbose=F)
```


```{r}
save(int.out, file = paste0(data_file_path, "crc_intout.RData"))
```

The output is a list containing

* estimated model parameters, common factor matrix `W`, factor loading matrices (per datasets) `H.list`, dataset specific scaling `lambda.list` and shift `b.list`
* convergence status `convergence`, number of iterations `niter`, and final objective function value `obj`, final objective delta `delta`
* parameters used for the function call, including `gamma`, `max.iter`, `tol`, `rep`

We obtain the integrated data via
```{r,  eval=T}
# nobs-by-nspecies expression matrix
exprs.int = do.call(rbind, int.out$H.list) %*% t(int.out$W)

# nobs-by-r low dimensiional representation
Hnorm = do.call(rbind, int.out$H.list) %*% diag(colSums(int.out$W))
```

```{r}
a = do.call(rbind, int.out$H.list)
b = diag(colSums(int.out$W))
```



Visualize the integrated data via UMAP plot,
```{r, eval=T, fig.align='center',fig.show="hold", fig.cap='Integrated pancreas islet cells data sests from four technologies.'}
studies = do.call(c, lapply(data.list$metadata.list[ref.data], function(x) x$study_name))
studies_all = do.call(c, lapply(data.list$metadata.list, function(x) x$study_name))
# celltype = do.call(c, data.list$labels.list[ref.data])
celltype = tech

umap.out = plot_umap(X=Hnorm, labels=studies, 
                     pca = NULL, n_components = 2, n_neighbors = 50, min_dist = 0.1, # umap parameters
                     point.size = 0.6, alpha=0.8, title=NULL, legend.name='study', # figure parameters
                     seed=42)

p1 = umap.out$p # colored by technologies
# p2 = plot_umap(labels=tech, point.size = 0.5, alpha=0.5, legend.name='technology', emb=umap.out$emb)$p # colored by cell types
p1
# p2
```


Calculate the alignment score for each study

```{r, eval=T}
alignment.score.out = calculate_alignment(X.list=list(Hnorm), k=30, balanced=F, dataset=studies, labels=studies)
alignment.score.out$alignment.per.dataset
```

## Data transfer

Next we transfer the learned factor matrix to the smallest data set, containing 60 observations from study GuptaA_2019.

```{r, eval=T}
# transfer: run time 10 seconds
tf.out = CFITTransfer(Xtarget=exprs.list[[2]], Wref=int.out$W, max.niter = 100, seed=0, verbose=F)
```

Visualize the transferred results via UMAP plot,
```{r, eval=T, fig.align='center',fig.show="hold", fig.cap='cFIT transfer results.'}
Hnorm = rbind(do.call(rbind, int.out$H.list), tf.out$H) %*% diag(colSums(int.out$W))
source = rep(c('reference','target'), c(nrow(do.call(rbind, int.out$H.list)), nrow(tf.out$H)))
celltype = do.call(c, c(data.list$labels.list[ref.data],data.list$labels.list[4]))

umap.out = plot_umap(X=Hnorm, labels=source, min_dist = 0.1, # umap parameters
                     point.size = 0.6, alpha=0.8, title=NULL, legend.name='source', 
                     cols=c('grey80','red'), seed=0)

p1 = umap.out$p # colored by source
# p2 = plot_umap(labels=celltype, point.size = 0.5, alpha=0.5, legend.name='cell type',  emb=umap.out$emb)$p # colored by cell types
p1
# p2
```


Assign labels for each cell in the target data by querying the cell type of target cells within k nearest neighbors.


```{r, eval=T}
est.labels = asign_labels(exprs.source=do.call(rbind, int.out$H.list), 
                           exprs.target=tf.out$H, 
                           labels.source=do.call(c, data.list$labels.list[ref.data]))
plotContTable(est_label=est.labels, true_label=data.list$labels.list[[2]], ylab='Mapped type')
```
```{r}
a = table(est.labels, data.list$labels.list[[2]])[1:2,1:2]
```


```{r}
# Confusion matrix
confusion_matrix <- confusionMatrix(table(est.labels, data.list$labels.list[[2]])[2:3,1:2])

# Precision, recall, F1-score
precision <- confusion_matrix$byClass["Precision"]
recall <- confusion_matrix$byClass["Recall"]
f1_score <- confusion_matrix$byClass["F1"]

# Print the metrics
print(precision)
print(recall)
print(f1_score)
```

